
## Dependencies and setup

```{r, message=FALSE}
library(tidyverse)
library(reshape2) # (Wickham, 2007)
library(cowplot)  # (Wilke, 2019)
```

We conducted these analyses using the following computing environment:

```{r}
print(version)
```

```{r}
# Build relative path from R's working directory to what is actually the working
# directory for this experiment.
experiment_slug <- "2025-10-29-arch-enumeration"
working_directory <- paste(
  "experiments",
  experiment_slug,
  "analysis",
  sep = "/"
)
```

```{r}
# Use the cowplot theme as our default for ggplots
theme_set(theme_cowplot())
# Create directory to save plots
plot_dir <- paste(
  working_directory,
  "plots",
  sep = "/"
)
dir.create(
  plot_dir,
  showWarnings = FALSE
)
```

## Load and preprocess data

```{r}
summary_data_path <- paste(
  working_directory,
  "data",
  "summary.csv",
  sep = "/"
)
summary_data <- read_csv(summary_data_path)

summary_data <- summary_data %>%
  mutate(
    BIT_FLIP_PROB = as.factor(BIT_FLIP_PROB),
    DRIFT = TOURNAMENT_SIZE == 1,
    gene_starts = as.factor(gene_starts),
    CHANGE_FREQUENCY = as.factor(CHANGE_FREQUENCY),
    CHANGE_MAGNITUDE = as.factor(CHANGE_MAGNITUDE),
    coding_sites_factor = as.factor(coding_sites),
    neutral_sites_factor = as.factor(neutral_sites),
    GRADIENT_MODEL = GRADIENT_MODEL == 1
  ) %>%
  mutate(
    chg_rate_label = case_when(
      DRIFT == TRUE ~ "drift",
      CHANGE_FREQUENCY == "0" ~ "0",
      .default = paste(CHANGE_MAGNITUDE, CHANGE_FREQUENCY, sep = "/")
    )
  ) %>%
  arrange(coding_sites)
```

Summarize distribution of coding sites, neutral sites across architectures.
Each treatment represents a different architecture, so we can group replicates together for this.

```{r}
architecture_summary <- summary_data %>%
  group_by(
    gene_starts
  ) %>%
  summarize(
    n = n(),
    same_coding_site_count = all(coding_sites == first(coding_sites)),
    same_neutral_site_count = all(neutral_sites == first(neutral_sites)),
    coding_sites = first(coding_sites),
    neutral_sites = first(neutral_sites)
  ) %>%
  ungroup() %>%
  mutate(
    coding_sites_factor = as.factor(coding_sites),
    neutral_sites_factor = as.factor(neutral_sites)
  )

```

```{r}
gradient_model_summary_data <- summary_data %>%
  filter(GRADIENT_MODEL == TRUE)
nk_model_summary_data <- summary_data %>%
  filter(GRADIENT_MODEL == FALSE)
```

## Genetic architecture characteristics

Here, we characterize the 65 meaningfully distinct architectures with 4, 4-bit genes and length-16 genomes.

### Coding sites

```{r}
dist_coding_sites_plot <- architecture_summary %>%
  ggplot(aes(x = coding_sites_factor)) +
  geom_histogram(stat = "count") +
  scale_y_continuous(
    breaks = seq(1, 11, 1),
    labels = seq(1, 11, 1)
  ) +
  labs(
    title = "Distribution of coding sites",
    x = "Coding sites",
    y = "Count"
  )

ggsave(
  file = paste(
    plot_dir,
    "dist_coding_sites_plot.pdf",
    sep = "/"
  ),
  plot = dist_coding_sites_plot,
  width = 10,
  height = 10
)
```

### Neutral sites

```{r}
dist_neutral_sites_plot <- architecture_summary %>%
  ggplot(aes(x = neutral_sites_factor)) +
  geom_histogram(stat = "count") +
  scale_y_continuous(
    breaks = seq(1, 11, 1),
    labels = seq(1, 11, 1)
  ) +
  labs(
    title = "Distribution of neutral sites",
    x = "Neutral sites",
    y = "Count"
  )

ggsave(
  file = paste(
    plot_dir,
    "dist_neutral_sites_plot.pdf",
    sep = "/"
  ),
  plot = dist_neutral_sites_plot,
  width = 10,
  height = 10
)
```

## Gradient model analysis

### Fitness

#### Maximum fitness across architectures

```{r}
gradient_max_fitness_by_arch_plot <- gradient_model_summary_data %>%
  ggplot(
    aes(
      x = gene_starts,
      y = fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping=aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "Gradient fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "gradient_max_fitness_by_arch_plot.pdf",
    sep = "/"
  ),
  plot = gradient_max_fitness_by_arch_plot,
  width = 30,
  height = 40
)
```

#### Average population fitness across architectures

```{r}
gradient_avg_fitness_by_arch_plot <- gradient_model_summary_data %>%
  ggplot(
    aes(
      x = gene_starts,
      y = mean_fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping = aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "Gradient fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "gradient_avg_fitness_by_arch_plot.pdf",
    sep = "/"
  ),
  plot = gradient_avg_fitness_by_arch_plot,
  width = 30,
  height = 40
)
```

#### Maximum fitness by coding site count

```{r}
gradient_max_fitness_by_coding_sites_plot <- gradient_model_summary_data %>%
  ggplot(
    aes(
      x = coding_sites_factor,
      y = fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping=aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "Gradient fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "gradient_max_fitness_by_coding_sites_plot.pdf",
    sep = "/"
  ),
  plot = gradient_max_fitness_by_coding_sites_plot,
  width = 10,
  height = 10
)
```

```{r}
gradient_model_summary_data %>%
  ggplot(
    aes(x = coding_sites, y = fitness)
  ) +
  geom_jitter(alpha = 0.95) +
  scale_fill_continuous(
    type = "viridis",
    trans = "log",
    name = "Count",
    breaks = c(1, 10, 100)
  ) +
  scale_x_continuous(
    name = "Coding Sites",
    limits = c(0, 17),
    breaks = seq(0, 16, 1)
  ) +
  ylab("Fitness") +
  facet_wrap(
    ~ BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  )
```

```{r}
data_gradient_bf3 <- filter(
  gradient_model_summary_data,
  BIT_FLIP_PROB == 0.3
)
data_gradient_bf03 <- filter(
  gradient_model_summary_data,
  BIT_FLIP_PROB == 0.03
)
data_gradient_bf003 <- filter(
  gradient_model_summary_data,
  BIT_FLIP_PROB == 0.003
)
cor.test(
  x = data_gradient_bf003$coding_sites,
  y = data_gradient_bf003$fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_gradient_bf03$coding_sites,
  y = data_gradient_bf03$fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_gradient_bf3$coding_sites,
  y = data_gradient_bf3$fitness,
  method = "spearman",
  exact = FALSE
)
```

#### Average fitness by coding site count

```{r}
gradient_avg_fitness_by_coding_sites_plot <- gradient_model_summary_data %>%
  ggplot(
    aes(
      x = coding_sites_factor,
      y = mean_fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping=aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "Gradient fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "gradient_avg_fitness_by_coding_sites_plot.pdf",
    sep = "/"
  ),
  plot = gradient_avg_fitness_by_coding_sites_plot,
  width = 10,
  height = 10
)
```

```{r}
gradient_model_summary_data %>%
  ggplot(
    aes(x = coding_sites, y = mean_fitness)
  ) +
  geom_jitter(alpha = 0.95) +
  scale_fill_continuous(
    type = "viridis",
    trans = "log",
    name = "Count",
    breaks = c(1, 10, 100)
  ) +
  scale_x_continuous(
    name = "Coding Sites",
    limits = c(0, 17),
    breaks = seq(0, 16, 1)
  ) +
  ylab("Mean fitness") +
  facet_wrap(
    ~ BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  )
```

```{r}
data_gradient_bf3 <- filter(
  gradient_model_summary_data,
  BIT_FLIP_PROB == 0.3
)
data_gradient_bf03 <- filter(
  gradient_model_summary_data,
  BIT_FLIP_PROB == 0.03
)
data_gradient_bf003 <- filter(
  gradient_model_summary_data,
  BIT_FLIP_PROB == 0.003
)
cor.test(
  x = data_gradient_bf003$coding_sites,
  y = data_gradient_bf003$mean_fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_gradient_bf03$coding_sites,
  y = data_gradient_bf03$mean_fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_gradient_bf3$coding_sites,
  y = data_gradient_bf3$mean_fitness,
  method = "spearman",
  exact = FALSE
)
```

## NK model analysis

### Fitness

#### Maximum fitness across architectures

```{r}
nk_max_fitness_by_arch_plot <- nk_model_summary_data %>%
  ggplot(
    aes(
      x = gene_starts,
      y = fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping=aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "nk fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "nk_max_fitness_by_arch_plot.pdf",
    sep = "/"
  ),
  plot = nk_max_fitness_by_arch_plot,
  width = 30,
  height = 40
)
```

#### Average population fitness across architectures

```{r}
nk_avg_fitness_by_arch_plot <- nk_model_summary_data %>%
  ggplot(
    aes(
      x = gene_starts,
      y = mean_fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping = aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "nk fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "nk_avg_fitness_by_arch_plot.pdf",
    sep = "/"
  ),
  plot = nk_avg_fitness_by_arch_plot,
  width = 30,
  height = 40
)
```

#### Maximum fitness by coding site count

```{r}
nk_max_fitness_by_coding_sites_plot <- nk_model_summary_data %>%
  ggplot(
    aes(
      x = coding_sites_factor,
      y = fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping=aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "nk fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "nk_max_fitness_by_coding_sites_plot.pdf",
    sep = "/"
  ),
  plot = nk_max_fitness_by_coding_sites_plot,
  width = 10,
  height = 10
)
```

```{r}
nk_model_summary_data %>%
  ggplot(
    aes(x = coding_sites, y = fitness)
  ) +
  geom_jitter(alpha = 0.95) +
  scale_fill_continuous(
    type = "viridis",
    trans = "log",
    name = "Count",
    breaks = c(1, 10, 100)
  ) +
  scale_x_continuous(
    name = "Coding Sites",
    limits = c(0, 17),
    breaks = seq(0, 16, 1)
  ) +
  ylab("Fitness") +
  facet_wrap(
    ~ BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  )
```

```{r}
data_nk_bf3 <- filter(
  nk_model_summary_data,
  BIT_FLIP_PROB == 0.3
)
data_nk_bf03 <- filter(
  nk_model_summary_data,
  BIT_FLIP_PROB == 0.03
)
data_nk_bf003 <- filter(
  nk_model_summary_data,
  BIT_FLIP_PROB == 0.003
)
cor.test(
  x = data_nk_bf003$coding_sites,
  y = data_nk_bf003$fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_nk_bf03$coding_sites,
  y = data_nk_bf03$fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_nk_bf3$coding_sites,
  y = data_nk_bf3$fitness,
  method = "spearman",
  exact = FALSE
)
```

#### Average fitness by coding site count

```{r}
nk_avg_fitness_by_coding_sites_plot <- nk_model_summary_data %>%
  ggplot(
    aes(
      x = coding_sites_factor,
      y = mean_fitness,
      fill = coding_sites_factor
    )
  ) +
  geom_point(
    mapping=aes(color = coding_sites_factor),
    position = position_jitter(width = .05),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  labs(
    title = "nk fitness model"
  ) +
  facet_wrap(
    ~BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  )

ggsave(
  file = paste(
    plot_dir,
    "nk_avg_fitness_by_coding_sites_plot.pdf",
    sep = "/"
  ),
  plot = nk_avg_fitness_by_coding_sites_plot,
  width = 10,
  height = 10
)
```

```{r}
nk_model_summary_data %>%
  ggplot(
    aes(x = coding_sites, y = mean_fitness)
  ) +
  geom_jitter(alpha = 0.95) +
  scale_fill_continuous(
    type = "viridis",
    trans = "log",
    name = "Count",
    breaks = c(1, 10, 100)
  ) +
  scale_x_continuous(
    name = "Coding Sites",
    limits = c(0, 17),
    breaks = seq(0, 16, 1)
  ) +
  ylab("Mean fitness") +
  facet_wrap(
    ~ BIT_FLIP_PROB,
    ncol = 1,
    labeller = label_both
  )
```

```{r}
cor.test(
  x = data_nk_bf003$coding_sites,
  y = data_nk_bf003$mean_fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_nk_bf03$coding_sites,
  y = data_nk_bf03$mean_fitness,
  method = "spearman",
  exact = FALSE
)
cor.test(
  x = data_nk_bf3$coding_sites,
  y = data_nk_bf3$mean_fitness,
  method = "spearman",
  exact = FALSE
)
```