## Dependencies and setup

```{r, message=FALSE}
library(tidyverse)
library(reshape2) # (Wickham, 2007)
library(cowplot)  # (Wilke, 2019)
library(hexbin)   # (Carr, Lewin-Kog, Maechler, and Sarkar, 2020)
library(knitr)
library(kableExtra)
```

We conducted these analyses using the following computing environment:

```{r}
print(version)
```

```{r}
# Build relative path from R's working directory to what is actually the working
# directory for this experiment.
experiment_slug <- "2025-12-22-mutation-rate-gradient"
working_directory <- paste(
  "experiments",
  experiment_slug,
  "analysis",
  sep = "/"
)
```

```{r}
# Use the cowplot theme as our default for ggplots
theme_set(theme_cowplot())
# Create directory to save plots
plot_dir <- paste(
  working_directory,
  "plots",
  sep = "/"
)
dir.create(
  plot_dir,
  showWarnings = FALSE
)
```

## Load and preprocess data

```{r}
sig_thresh <- 0.5
```

```{r}
summary_data_path <- paste(
  working_directory,
  "data",
  "summary.csv",
  sep = "/"
)
summary_data <- read_csv(summary_data_path)
summary_data <- summary_data %>%
  mutate(
    BIT_FLIP_PROB = as.factor(BIT_FLIP_PROB),
    DRIFT = TOURNAMENT_SIZE == 1
  ) %>%
  mutate(
    chg_rate_label = case_when(
      DRIFT ~ "Drift",
      CHANGE_MAGNITUDE == "0" ~ "Static",
      CHANGE_FREQUENCY == "0" ~ "Static",
      .default = paste(
        CHANGE_MAGNITUDE,
        CHANGE_FREQUENCY,
        sep = "/"
      )
    )
  ) %>%
  mutate(
    chg_rate_label = factor(
      chg_rate_label,
      levels = c(
        "Static",
        "1/1024",
        "1/512",
        "1/256",
        "1/128",
        "1/64",
        "1/32",
        "1/16",
        "1/8",
        "1/4",
        "1/2",
        "1/1",
        "2/1",
        "4/1",
        "8/1",
        "16/1",
        "32/1",
        "64/1",
        "128/1",
        "256/1",
        "512/1",
        "1024/1",
        "2048/1",
        "4096/1",
        "Drift"
      )
    )
  )

phase0_summary_data <- summary_data %>%
  filter(
    evo_phase == 0 & update == 50000
  )
phase1_summary_data <- summary_data %>%
  filter(evo_phase == 1 & update == 60000)
```

## Evolution experiment

We evolved populations for 50,000 generations under a range of environmental change rates.

### Fitnesses

Note issue with fitnesses for changing environments.

#### Gradient Fitness Model

```{r}
grad_final_fitness_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = fitness,
      color = chg_rate_label
    )
  ) +
  geom_boxplot(
    position = "dodge"
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Fitness (of best organism)"
  ) +
  theme()

ggsave(
  file = paste(
    plot_dir,
    "grad_final_fitness_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_fitness_plt,
  width = 10,
  height = 10
)
```

### Coding sites

#### Gradient fitness model

```{r}
grad_final_coding_sites_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = coding_sites,
      color = chg_rate_label
    )
  ) +
  geom_boxplot(
    position = "dodge"
  ) +
  scale_y_continuous(
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Coding sites (in best organism)"
  ) +
  theme()

ggsave(
  file = paste(
    plot_dir,
    "grad_final_coding_sites_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_coding_sites_plt,
  width = 10,
  height = 10
)
```

### Genome Length

#### Gradient fitness model

```{r}
grad_final_genome_length_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = genome_length,
      color = chg_rate_label
    )
  ) +
  geom_boxplot(
    position = "dodge"
  ) +
  scale_y_continuous(
    limits = c(0, 1025)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Genome length (of best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_final_genome_length_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_genome_length_plt,
  width = 10,
  height = 10
)
```


### Neutral sites

#### Gradient fitness model

```{r}
grad_final_neutral_sites_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = neutral_sites,
      color = chg_rate_label
    )
  ) +
  geom_boxplot() +
  scale_y_continuous(
    limits = c(0, 1025)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Neutral sites (of best organism)"
  ) +
  theme()

ggsave(
  file = paste(
    plot_dir,
    "grad_final_neutral_sites_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_neutral_sites_plt,
  width = 10,
  height = 10
)
```

## Transfer experiments

We lock down genetic architectures and evolve populations for an additional 10,000
generations in a new (random) static environment.

Are architectures evolved under changing environments more evolvable than those
evolved under static environments?

### Fitness

#### Gradient fitness model

```{r}
grad_second_phase_fitness_plt <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = fitness,
      color = chg_rate_label
    )
  ) +
  geom_boxplot(position="dodge") +
  labs(
    title = "Gradient Fitness Model - Second experiment phase",
    x = "Mutation rate",
    y = "Final fitness (of best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_second_phase_fitness_plt.pdf",
    sep = "/"
  ),
  plot = grad_second_phase_fitness_plt,
  width = 10,
  height = 10
)
```

### Fitness vs. Number of coding sites

#### Gradient fitness Model

```{r}
grad_second_phase_fitness_v_coding_plt <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = coding_sites,
      y = fitness,
      color = BIT_FLIP_PROB
    )
  ) +
  geom_point() +
  labs(
    title = "Gradient Fitness Model - Second experiment phase",
    x = "Coding sites",
    y = "Fitness"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_second_phase_fitness_v_coding_plt.pdf",
    sep = "/"
  ),
  plot = grad_second_phase_fitness_v_coding_plt,
  width = 10,
  height = 10
)
```

## Manuscript

Coding sites, just static

```{r}
grad_final_coding_sites_plt_manu <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1 & chg_rate_label == "Static") %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = coding_sites
    )
  ) +
  geom_boxplot(
    position = "dodge"
  ) +
  scale_y_continuous(
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  labs(
    # title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Coding sites\n(in most fit genotype)"
  ) +
  theme(
    legend.position = "none",
    axis.title.y = element_text(
      size = 12
    ),
    axis.text.y = element_text(
      size = 10
    ),
    axis.title.x = element_text(size = 12),
    axis.text.x = element_text(
      size = 8
    )
  )

ggsave(
  file = paste(
    plot_dir,
    "grad_final_coding_sites_plt_manu.png",
    sep = "/"
  ),
  plot = grad_final_coding_sites_plt_manu,
  width = 7,
  height = 4,
  dpi = 600
)
```

```{r}
d <- filter(
  phase0_summary_data,
  GRADIENT_MODEL == 1 & chg_rate_label == "Static"
)
kt <- kruskal.test(
  formula = coding_sites ~ BIT_FLIP_PROB,
  data = d
)
kt
if (kt$p.value <= sig_thresh) {
  wt <- pairwise.wilcox.test(
    x = d$coding_sites,
    g = d$BIT_FLIP_PROB,
    exact = FALSE,
    p.adjust.method = "bonferroni"
  )
  wt
}
```

Fitness of static

```{r}
fitness_summary <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1 & chg_rate_label == "Static") %>%
  group_by(
    BIT_FLIP_PROB
  ) %>%
  summarize(
    avg_fitness = mean(fitness),
    median_fitness = median(fitness),
    variance_fitness = var(fitness),
    min_fitness = min(fitness),
    max_fitness = max(fitness)
  ) %>%
  arrange(desc(median_fitness))

# coding_sites_summary$min_coding_sites
# coding_sites_summary$max_coding_sites
```

```{r}
d <- filter(
  phase0_summary_data,
  GRADIENT_MODEL == 1 & chg_rate_label == "Drift"
)
kt <- kruskal.test(
  formula = coding_sites ~ BIT_FLIP_PROB,
  data = d
)
kt
```

<!-- ```{r}
cor_data <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1)
cor.test(
  x = cor_data$coding_sites,
  y = cor_data$fitness,
  method = "spearman",
  exact = FALSE
)
``` -->
