## Dependencies and setup

```{r, message=FALSE}
library(tidyverse)
library(reshape2) # (Wickham, 2007)
library(cowplot)  # (Wilke, 2019)
library(hexbin)   # (Carr, Lewin-Kog, Maechler, and Sarkar, 2020)
library(knitr)
library(kableExtra)
library(khroma)
```

We conducted these analyses using the following computing environment:

```{r}
print(version)
```

```{r}
# Build relative path from R's working directory to what is actually the working
# directory for this experiment.
experiment_slug <- "2026-01-06-env-chg-and-mutation"
working_directory <- paste(
  "experiments",
  experiment_slug,
  "analysis",
  sep = "/"
)
```

```{r}
# Use the cowplot theme as our default for ggplots
theme_set(theme_cowplot())
# Create directory to save plots
plot_dir <- paste(
  working_directory,
  "plots",
  sep = "/"
)
dir.create(
  plot_dir,
  showWarnings = FALSE
)
```

## Load and preprocess data

```{r}
sig_thresh <- 0.5
```

```{r}
summary_data_path <- paste(
  working_directory,
  "data",
  "summary.csv",
  sep = "/"
)
summary_data <- read_csv(summary_data_path)
summary_data <- summary_data %>%
  mutate(
    BIT_FLIP_PROB = as.factor(BIT_FLIP_PROB),
    DRIFT = TOURNAMENT_SIZE == 1
  ) %>%
  mutate(
    chg_rate_label = case_when(
      DRIFT ~ "Drift",
      CHANGE_MAGNITUDE == "0" ~ "Static",
      CHANGE_FREQUENCY == "0" ~ "Static",
      .default = paste(
        CHANGE_MAGNITUDE,
        CHANGE_FREQUENCY,
        sep = "/"
      )
    )
  ) %>%
  mutate(
    chg_rate_label = factor(
      chg_rate_label,
      levels = c(
        "Static",
        "1/1024",
        "1/512",
        "1/256",
        "1/128",
        "1/64",
        "1/32",
        "1/16",
        "1/8",
        "1/4",
        "1/2",
        "1/1",
        "2/1",
        "4/1",
        "8/1",
        "16/1",
        "32/1",
        "64/1",
        "128/1",
        "256/1",
        "512/1",
        "1024/1",
        "2048/1",
        "4096/1",
        "Drift"
      )
    )
  )

phase0_summary_data <- summary_data %>%
  filter(
    evo_phase == 0 & update == 50000
  )
```

## Evolution experiment

We evolved populations for 50,000 generations under a range of environmental change rates.

### Fitnesses

Note issue with fitnesses for changing environments.

#### Gradient Fitness Model

```{r}
grad_final_fitness_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = fitness,
      color = chg_rate_label
    )
  ) +
  geom_boxplot(
    position = "dodge"
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Fitness (of best organism)"
  ) +
  facet_wrap(~chg_rate_label) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

ggsave(
  file = paste(
    plot_dir,
    "grad_final_fitness_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_fitness_plt,
  width = 20,
  height = 20
)
```

### Coding sites

#### Gradient fitness model

```{r}
grad_final_coding_sites_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = coding_sites,
      color = chg_rate_label
    )
  ) +
  geom_boxplot(
    position = "dodge"
  ) +
  scale_y_continuous(
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Mutation rate",
    y = "Coding sites (in best organism)"
  ) +
  facet_wrap(~chg_rate_label) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

ggsave(
  file = paste(
    plot_dir,
    "grad_final_coding_sites_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_coding_sites_plt,
  width = 20,
  height = 20
)
```

## Manuscript



```{r}
coding_sites_summary <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  group_by(
    BIT_FLIP_PROB,
    chg_rate_label
  ) %>%
  summarize(
    median_coding_sites = median(coding_sites)
  )

grad_chg_rate_vs_mut_plt_manu <- coding_sites_summary %>%
  ggplot(
    aes(
      x = BIT_FLIP_PROB,
      y = chg_rate_label,
      fill = median_coding_sites
    )
  ) +
  geom_tile() +
  scale_fill_lajolla(
    reverse = TRUE
  ) +
  labs(
    x = "Mutation rate",
    y = "Environmental change",
    fill = "Coding sites"
  ) +
  theme(
    legend.position = "right",
    axis.title.y = element_text(size = 14),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 11)
  )

ggsave(
  file = paste(
    plot_dir,
    "grad_chg_rate_vs_mut_plt_manu.pdf",
    sep = "/"
  ),
  plot = grad_chg_rate_vs_mut_plt_manu,
  width = 8,
  height = 7,
  dpi = 600
)
```

```{r}
# interaction <- aov(
#   coding_sites ~ BIT_FLIP_PROB * chg_rate_label,
#   data = filter(phase0_summary_data, GRADIENT_MODEL == 1)
# )

d <- phase0_summary_data %>% filter(GRADIENT_MODEL == 1)
mod_chg <- lm(coding_sites ~ chg_rate_label, data = d)
mod_mut <- lm(coding_sites ~ BIT_FLIP_PROB, data = d)
mod_chgxmut <- lm(coding_sites ~ chg_rate_label * BIT_FLIP_PROB, data = d)
mod_chg_and_mut <- lm(coding_sites ~ chg_rate_label + BIT_FLIP_PROB, data = d)
anova(
  mod_chgxmut,
  mod_chg_and_mut
)
```

<!-- ```{r}
cor_data <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1)
cor.test(
  x = cor_data$coding_sites,
  y = cor_data$fitness,
  method = "spearman",
  exact = FALSE
)
``` -->
