## Dependencies and setup

```{r, message=FALSE}
library(tidyverse)
library(reshape2) # (Wickham, 2007)
library(cowplot)  # (Wilke, 2019)
library(hexbin)   # (Carr, Lewin-Kog, Maechler, and Sarkar, 2020)
library(knitr)
library(kableExtra)
```

We conducted these analyses using the following computing environment:

```{r}
print(version)
```

```{r}
# Build relative path from R's working directory to what is actually the working
# directory for this experiment.
experiment_slug <- "2025-12-18-env-chg-wider-gradient"
working_directory <- paste(
  "experiments",
  experiment_slug,
  "analysis",
  sep = "/"
)
```

```{r}
# Use the cowplot theme as our default for ggplots
theme_set(theme_cowplot())
# Create directory to save plots
plot_dir <- paste(
  working_directory,
  "plots",
  sep = "/"
)
dir.create(
  plot_dir,
  showWarnings = FALSE
)
```

## Load and preprocess data

```{r}
sig_thresh <- 0.5
```

```{r}
summary_data_path <- paste(
  working_directory,
  "data",
  "summary.csv",
  sep = "/"
)
summary_data <- read_csv(summary_data_path)
summary_data <- summary_data %>%
  mutate(
    BIT_FLIP_PROB = as.factor(BIT_FLIP_PROB),
    DRIFT = TOURNAMENT_SIZE == 1
  ) %>%
  mutate(
    chg_rate_label = case_when(
      DRIFT ~ "Drift",
      CHANGE_MAGNITUDE == "0" ~ "Static",
      CHANGE_FREQUENCY == "0" ~ "Static",
      .default = paste(
        CHANGE_MAGNITUDE,
        CHANGE_FREQUENCY,
        sep = "/"
      )
    )
  ) %>%
  mutate(
    chg_rate_label = factor(
      chg_rate_label,
      levels = c(
        "Static",
        "1/1024",
        "1/512",
        "1/256",
        "1/128",
        "1/64",
        "1/32",
        "1/16",
        "1/8",
        "1/4",
        "1/2",
        "1/1",
        "2/1",
        "4/1",
        "8/1",
        "16/1",
        "32/1",
        "64/1",
        "128/1",
        "256/1",
        "512/1",
        "1024/1",
        "2048/1",
        "4096/1",
        "Drift"
      )
    )
  )

phase0_summary_data <- summary_data %>%
  filter(
    evo_phase == 0 & update == 50000
  )
phase1_summary_data <- summary_data %>%
  filter(evo_phase == 1 & update == 60000)
```

## Evolution experiment

We evolved populations for 50,000 generations under a range of environmental change rates.

### Fitnesses

Note issue with fitnesses for changing environments.

#### Gradient Fitness Model

```{r}
grad_final_fitness_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = fitness,
      color = chg_rate_label
    )
  ) +
  geom_boxplot() +
  labs(
    title = "Gradient Fitness Model",
    x = "Environmental Change",
    y = "Fitness (of best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_final_fitness_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_fitness_plt,
  width = 10,
  height = 10
)
```

### Coding sites

#### Gradient fitness model

```{r}
grad_final_coding_sites_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = coding_sites,
      color = chg_rate_label
    )
  ) +
  geom_boxplot() +
  scale_y_continuous(
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Environmental Change",
    y = "Coding sites (in best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_final_coding_sites_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_coding_sites_plt,
  width = 10,
  height = 10
)
```

Statistical comparison of coding sites across treatments.

```{r}
d <- filter(phase0_summary_data, GRADIENT_MODEL == 1)
kt <- kruskal.test(
  formula = coding_sites ~ chg_rate_label,
  data = d
)
kt
if (kt$p.value <= sig_thresh) {
  wt <- pairwise.wilcox.test(
    x = d$coding_sites,
    g = d$chg_rate_label,
    exact = FALSE,
    p.adjust.method = "bonferroni"
  )
  wt
  # stats_table <- kbl(wt) %>%
  #   kable_styling(latex_options = "striped")
  # save_kable(
  #   extinction_table,
  #   paste(
  #     plot_dir,
  #     "extinctions.pdf",
  #     sep = "/"
  #   )
  # )
}
```

TODO - Calculate effect sizes

```{r}
coding_sites_summary <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  group_by(chg_rate_label) %>%
  summarize(
    avg_coding_sites = mean(coding_sites),
    median_coding_sites = median(coding_sites),
    variance_coding_sites = var(coding_sites),
    min_coding_sites = min(coding_sites),
    max_coding_sites = max(coding_sites)
  ) %>%
  arrange(desc(median_coding_sites))

# coding_sites_summary$min_coding_sites
# coding_sites_summary$max_coding_sites
```

### Coding sites vs first-phase fitness

Are more compact genomes less fit at the end of phase 1?

```{r}
coding_sites_vs_phase0_fitness_plt <- phase0_summary_data %>%
  filter(
    GRADIENT_MODEL == 1 & chg_rate_label == "Static"
  ) %>%
  ggplot(
    aes(
      x = coding_sites,
      y = fitness
    )
  ) +
  geom_point(alpha = 0.95) +
  scale_x_continuous(
    name = "Coding Sites",
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  scale_fill_continuous(
    type = "viridis",
    trans = "log",
    name = "Count",
    breaks = c(1, 10, 100)
  ) +
  labs(
    x = "Coding Sites",
    y = "Phase 0 fitness"
  )

ggsave(
  file = paste(
    plot_dir,
    "coding_sites_vs_phase0_fitness_plt.pdf",
    sep = "/"
  ),
  plot = coding_sites_vs_phase0_fitness_plt,
  width = 10,
  height = 10
)
```
```{r}
static_phase0 <- phase0_summary_data %>%
  filter(
    GRADIENT_MODEL == 1 & chg_rate_label == "Static"
  )

phase0_summary_data %>%
  filter(
    GRADIENT_MODEL == 1
  ) %>%
  group_by(chg_rate_label) %>%
  summarize(
    avg_fitness = mean(fitness),
    min_fitness = min(fitness),
    max_fitness = max(fitness)
  ) %>%
  arrange(desc(avg_fitness))
```

### Genome Length

#### Gradient fitness model

```{r}
grad_final_genome_length_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = genome_length,
      color = chg_rate_label
    )
  ) +
  geom_boxplot() +
  scale_y_continuous(
    limits = c(0, 1025)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Environmental Change",
    y = "Genome length (of best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_final_genome_length_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_genome_length_plt,
  width = 10,
  height = 10
)
```

Statistical comparison of coding sites across treatments.

```{r}
d <- filter(phase0_summary_data, GRADIENT_MODEL == 1)
kt <- kruskal.test(
  formula = genome_length ~ chg_rate_label,
  data = d
)
kt
if (kt$p.value <= sig_thresh) {
  wt <- pairwise.wilcox.test(
    x = d$genome_length,
    g = d$chg_rate_label,
    exact = FALSE,
    p.adjust.method = "bonferroni"
  )
  wt
}
```


### Neutral sites

#### Gradient fitness model

```{r}
grad_final_neutral_sites_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = neutral_sites,
      color = chg_rate_label
    )
  ) +
  geom_boxplot() +
  scale_y_continuous(
    limits = c(0, 1025)
  ) +
  labs(
    title = "Gradient Fitness Model",
    x = "Environmental Change",
    y = "Neutral sites (of best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_final_neutral_sites_plt.pdf",
    sep = "/"
  ),
  plot = grad_final_neutral_sites_plt,
  width = 10,
  height = 10
)
```

```{r}
d <- filter(phase0_summary_data, GRADIENT_MODEL == 1)
kt <- kruskal.test(
  formula = neutral_sites ~ chg_rate_label,
  data = d
)
kt
if (kt$p.value <= sig_thresh) {
  wt <- pairwise.wilcox.test(
    x = d$neutral_sites,
    g = d$chg_rate_label,
    exact = FALSE,
    p.adjust.method = "bonferroni"
  )
  wt
}
```

## Transfer experiments

We lock down genetic architectures and evolve populations for an additional 10,000
generations in a new (random) static environment.

Are architectures evolved under changing environments more evolvable than those
evolved under static environments?

### Fitness

#### Gradient fitness model

```{r}
grad_second_phase_fitness_plt <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = fitness,
      color = chg_rate_label
    )
  ) +
  geom_boxplot() +
  labs(
    title = "Gradient Fitness Model - Second experiment phase",
    x = "Environmental Change",
    y = "Final fitness (of best organism)"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_second_phase_fitness_plt.pdf",
    sep = "/"
  ),
  plot = grad_second_phase_fitness_plt,
  width = 10,
  height = 10
)
```

```{r}
d <- filter(phase1_summary_data, GRADIENT_MODEL == 1)
kt <- kruskal.test(
  formula = fitness ~ chg_rate_label,
  data = d
)
kt
if (kt$p.value <= sig_thresh) {
  wt <- pairwise.wilcox.test(
    x = d$fitness,
    g = d$chg_rate_label,
    exact = FALSE,
    p.adjust.method = "bonferroni"
  )
  wt
}
```

### Fitness vs. Number of coding sites

#### Gradient fitness Model

```{r}
grad_second_phase_fitness_v_coding_plt <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = coding_sites,
      y = fitness,
      color = chg_rate_label
    )
  ) +
  geom_point() +
  labs(
    title = "Gradient Fitness Model - Second experiment phase",
    x = "Coding sites",
    y = "Fitness"
  ) +
  theme(legend.position = "none")

ggsave(
  file = paste(
    plot_dir,
    "grad_second_phase_fitness_v_coding_plt.pdf",
    sep = "/"
  ),
  plot = grad_second_phase_fitness_v_coding_plt,
  width = 10,
  height = 10
)
```

```{r}
cor_data <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1)
cor.test(
  x = cor_data$coding_sites,
  y = cor_data$fitness,
  method = "spearman",
  exact = FALSE
)
```

## Publication figures

### Changing environments promote modularity

```{r}

static_median <- median(
  filter(
    phase0_summary_data,
    GRADIENT_MODEL == 1 & chg_rate_label == "Static"
  )$coding_sites
)
chg_env_promote_modularity_plt <- phase0_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = coding_sites
    )
  ) +
  geom_boxplot(
    # width = 0.1
    # outlier.shape = NA,
    # alpha = 0.5
  ) +
  # geom_point(
  #   position = position_jitter(width = .15),
  #   size = .5,
  #   alpha = 0.8
  # ) +
  geom_hline(
    yintercept = static_median,
    linetype = "dashed",
    color = "gray",
    alpha = 0.9
  ) +
  labs(
    x = "Environmental Change",
    y = "Coding Sites\n(in most fit genotype)"
  ) +
  scale_y_continuous(
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  theme(
    legend.position = "none",
    axis.title.y = element_text(
      size = 12
    ),
    axis.text.y = element_text(
      size = 10
    ),
    axis.title.x = element_text(size = 12),
    axis.text.x = element_text(size = 8)
  )

ggsave(
  file = paste(
    plot_dir,
    "chg_env_promote_modularity_plt.png",
    sep = "/"
  ),
  plot = chg_env_promote_modularity_plt,
  width = 8,
  height = 4,
  dpi = 600
)
```

### Modularity promotes evolvability

```{r}
modularity_promote_evolvability_plt <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = chg_rate_label,
      y = fitness
    )
  ) +
  geom_boxplot(
    # width = 0.1
    # outlier.shape = NA,
    # alpha = 0.5
  ) +
  # geom_point(
  #   position = position_jitter(width = .15),
  #   size = .5,
  #   alpha = 0.8
  # ) +
  labs(
    x = "Original environment (pre-transfer)",
    y = "Fitness achieved\nafter environment transfer"
  ) +
  theme(
    legend.position = "none",
    axis.title.y = element_text(
      size = 12
    ),
    axis.text.y = element_text(
      size = 10
    ),
    axis.title.x = element_text(size = 12),
    axis.text.x = element_text(
      size = 8,
      angle = 30,
      hjust = 1
    )
  )

ggsave(
  file = paste(
    plot_dir,
    "modularity_promote_evolvability_plt.png",
    sep = "/"
  ),
  plot = modularity_promote_evolvability_plt,
  width = 7,
  height = 4,
  dpi = 600
)
```

```{r}
coding_sites_vs_fitness_manuscript_plt <- phase1_summary_data %>%
  filter(GRADIENT_MODEL == 1) %>%
  ggplot(
    aes(
      x = coding_sites,
      y = fitness
    )
  ) +
  geom_hex(alpha = 0.95) +
  scale_x_continuous(
    name = "Coding Sites",
    limits = c(0, 130),
    breaks = seq(0, 130, 20)
  ) +
  scale_fill_continuous(
    type = "viridis",
    trans = "log",
    name = "Count",
    breaks = c(1, 10, 100)
  ) +
  labs(
    x = "Coding Sites",
    y = "Fitness achieved\nafter environment transfer"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.title.y = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 12),
    axis.text.x = element_text(size = 8)
  )

ggsave(
  file = paste(
    plot_dir,
    "coding_sites_vs_fitness_manuscript_plt.png",
    sep = "/"
  ),
  plot = coding_sites_vs_fitness_manuscript_plt,
  width = 7,
  height = 4,
  dpi = 600
)
```

Combined

```{r}
y_lims <- c(8, 16.2)
plt_grid <- plot_grid(
  modularity_promote_evolvability_plt +
    scale_y_continuous(limits = y_lims) +
    theme(
      legend.position = "none",
      axis.title.y = element_text(size = 10),
      axis.text.y = element_text(size = 8),
      axis.title.x = element_text(size = 10),
      axis.text.x = element_text(size = 8, angle = 30, hjust = 1)
    ),
  coding_sites_vs_fitness_manuscript_plt +
    scale_y_continuous(limits = y_lims) +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 10),
      legend.text = element_text(size = 8),
      axis.title.y = element_text(size = 10),
      axis.text.y = element_text(size = 8),
      axis.title.x = element_text(size = 10),
      axis.text.x = element_text(size = 8)
    ),
  nrow = 1,
  ncol = 2,
  align = "h",
  labels = c("A)", "B)"),
  rel_widths = c(0.6, 0.4)
)
# plt_grid
save_plot(
  filename = paste(
    plot_dir,
    "modularity-promotes-evolvability-fig.pdf",
    sep = "/"
  ),
  plot = plt_grid,
  base_width = 8,
  base_height = 3,
  dpi = 600
)
```
